Geometric models reveal behavioral and neural signatures of transforming naturalistic experiences into episodic memories

### 2020-09-04 paper/main.tex
edits through end of main

Our general approach draws inspiration from prior work aimed at elucidating the neural and behavioral underpinnings of how we process dynamic naturalistic experiences and remember them later.  Our approach to identifying neural responses to naturalistic stimuli (including experiences) entails building an explicit model of the stimulus dynamics and searching for brain regions whose responses are consistent with the model~\citep[also see][]{HuthEtal12, HuthEtal16}.  In prior work, a series of studies from Uri Hasson's group~\citep{LernEtal11, SimoEtal16, ChenEtal17, BaldEtal17, ZadbEtal17} have developed a clever alternative approach: rather than building an explicit stimulus model, these studies instead search for brain responses (while experiencing the stimulus) that are reliably similar across individuals.  So called \textit{inter-subject correlation} (ISC) and \textit{inter-subject functional connectivity} (ISFC) analyses effectively treat other people's brain responses to the stimulus as a ``model'' of how its features change over time~\citep[also see][]{SimoChan20}.  These purely brain-driven approaches are well-suited to identifying which brain structures exhibit similar stimulus-driven responses across individuals.  Further, because neural response dynamics are observed data (rather than model approximations), such approaches do not require a detailed understanding of which stimulus properties or features might be driving the observed responses.  However, this also means that the specific stimulus features driving those responses are typically opaque to the researcher.  Our approach is complementary.  By explicitly modeling the stimulus dynamics, we are able to relate specific stimulus features to behavioral and neural dynamics.  However, when our model fails to accurately capture the stimulus dynamics that are truly driving behavioral and neural responses, our approach necessarily yields an incomplete characterization of the neural basis of the processes we are studying.

@@

The video stimulus was divided into 1,000 fine-grained ``scenes'' and annotated by an independent coder.  For each of these 1,000 scenes, the following information was recorded: a brief narrative description of what was happening, the location where the scene took place, whether that location was indoors or outdoors, the names of all characters on-screen, the name(s) of the character(s) in focus in the shot, the name(s) of the character(s) currently speaking, the camera angle of the shot, a transcription of any text appearing on-screen, and whether or not there was music present in the background.  Each scene was also tagged with its onset and offset time, in both seconds and TRs.

@@

\subsection*{Modeling the dynamic content of the episode and recall transcripts}
\subsubsection*{Topic modeling}
The input to the topic model we trained to characterize the dynamic content of the episode comprised 998 hand-generated annotations of short (mean: 2.96s) scenes spanning the video clip~(\citealp{ChenEtal17} generated 1000 annotations total; we removed two annotations referring to a break between the first and second scan sessions, during which no fMRI data were collected).  We concatenated the text for all of the annotated features within each segment, creating a ``bag of words'' describing each scene and performed some minor preprocessing (e.g., stemming possessive nouns and removing punctuation).  We then re-organized the text descriptions into overlapping sliding windows spanning (up to) 50 scenes each.  In other words, we estimated the ``context'' for each scene using the text descriptions of the preceding 25 scenes, the present scene, and the following 24 scenes.  To model the context for scenes near the beginning of the episode (i.e., within 25 scenes of the beginning or end), we created overlapping sliding windows that grew in size from one scene to the full length.  We also tapered the sliding window lengths at the end of the episode, whereby scenes within fewer than 24 scenes of the end of the episode were assigned sliding windows that extended to the end of the episode.  This procedure ensured that each scene's content was represented in the text corpus an equal number of times.
% ------ NOTE: ------
% need to refine explanation here a little bit
% - Paxton

We trained our model using these overlapping text samples with \texttt{scikit-learn}~\citep[version 0.19.1; ][]{PedrEtal11}, called from our high-dimensional visualization and text analysis software, \texttt{HyperTools}~\citep{HeusEtal18a}.  Specifically, we used the \texttt{CountVectorizer} class to transform the text from each window into a vector of word counts (using the union of all words across all scenes as the ``vocabulary,'' excluding English stop words); this yielded a number-of-windows by number-of-words \textit{word count} matrix.  We then used the \texttt{LatentDirichletAllocation} class (topics=100, method=`batch') to fit a topic model~\citep{BleiEtal03} to the word count matrix, yielding a number-of-windows (1047) by number-of-topics (100) \textit{topic proportions} matrix.  The topic proportions matrix describes the gradually evolving mix of topics (latent themes) present in each scene.  Next, we transformed the topic proportions matrix to match the 1976 fMRI volume acquisition times.  We assigned each topic vector to the timepoint (in seconds) midway between the beginning of the first scene and the end of the last scene in its corresponding sliding text window.  By doing so, we warped the linear temporal distance between consecutive topic vectors to align with the inconsistent temporal distance between consecutive annotations (whose durations varied greatly).  We then rescaled these timepoints to 1.5s TR units, and used linear interpolation to estimate a topic vector for each TR.  This resulted in a number-of-TRs (1976) by number-of-topics (100) matrix.

We created similar topic proportions matrices using hand-annotated transcripts of each participant's verbal recall of the episode~\citep[annotated by][]{ChenEtal17}.  We tokenized the transcript into a list of sentences, and then re-organized the list into overlapping sliding windows spanning (up to) 10 sentences each, analogously to how we parsed the episode annotations.  In turn, we transformed each window's sentences into a word count vector (using the same vocabulary as for the episode model), and then we used the topic model already trained on the episode scenes to compute the most probable topic proportions for each sliding window.  This yielded a number-of-windows (range: 83--312) by number-of-topics (100) topic proportions matrix for each participant.  These reflected the dynamic content of each participant's recalls.  Note: for details on how we selected the episode and recall window lengths and number of topics, see \textit{Supporting Information} and Figure~\topicopt.

@@

\paragraph{Probability of first recall (PFR).}  PFR curves~\citep{WelcBurn24, PostPhil65, AtkiShif68} reflect the probability that an item will be recalled first as a function of its serial position during encoding. To carry out this analysis, we initialized a number-of-participants (17) by number-of-episode-events (30) matrix of zeros. Then for each participant, we found the index of the episode event that was recalled first (i.e., the episode event whose topic vector was most strongly correlated with that of the first recall event) and filled in that index in the matrix with a 1.  Finally, we averaged over the rows of the matrix, resulting in a 1 by 30 array representing the proportion of participants that recalled an event first, as a function of the order of the event's appearance in the episode (Fig.~\ref{fig:list-learning}A).
% ------ NOTE: ------
% reiterate meaning of error ribbons in list-learning figure? (already noted in figure caption)
% - Paxton

@@

\paragraph{Serial position curve (SPC).} SPCs~\citep{Murd62a} reflect the proportion of participants that remember each item as a function of the items' serial positions during encoding. We initialized a number-of-participants (17) by number-of-episode-events (30) matrix of zeros. Then, for each recalled event, for each participant, we found the index of the episode event that the recalled event most closely matched (via the correlation between the events' topic vectors) and entered a 1 into that position in the matrix. This resulted in a matrix whose entries indicated whether or not each event was recalled by each participant (depending on whether the corresponding entires were set to one or zero).  Finally, we averaged over the rows of the matrix to yield a 1 by 30 array representing the proportion of participants that recalled each event as a function of the events' order appearance in the episode (Fig.~\ref{fig:list-learning}C).

@@

\subsubsection*{Visualizing the episode and recall topic trajectories}
We used the UMAP algorithm~\citep{McInEtal18} to project the 100-dimensional topic space onto a two-dimensional space for visualization (Figs.~\ref{fig:trajectory}, \ref{fig:topics}).  To ensure that all of the trajectories were projected onto the \textit{same} lower dimensional space, we computed the low-dimensional embedding on a ``stacked'' matrix created by vertically concatenating the events-by-topics topic proportions matrices for the episode, across-participants average recall and all 17 individual participants' recalls.  We then separated the rows of the result (a total-number-of-events by two matrix) back into individual matrices for the episode topic trajectory, across-participant average recall trajectory and the trajectories for each individual participant's recalls (Fig.~\ref{fig:trajectory}).  This general approach for discovering a shared low-dimensional embedding for a collections of high-dimensional observations follows \cite{HeusEtal18a}.

@@

\subsubsection*{Estimating the consistency of flow through topic space across participants}
In Figure~\ref{fig:trajectory}B, we present an analysis aimed at characterizing locations in topic space that different participants move through in a consistent way (via their recall topic trajectories).  The two-dimensional topic space used in our visualizations (Fig.~\ref{fig:trajectory}) comprised a $60 \times 60$ (arbitrary units) square.  We tiled this space with a $50 \times 50$ grid of evenly spaced vertices, and defined a circular area centered on each vertex whose radius was two times the distance between adjacent vertices (i.e., 2.4 units).  For each vertex, we examined the set of line segments formed by connecting each pair successively recalled events, across all participants, that passed through this circle.  We computed the distribution of angles formed by those segments and the $x$-axis, and used a Rayleigh test to determine whether the distribution of angles was reliably ``peaked'' (i.e., consistent across all transitions that passed through that local portion of topic space).  To create Figure~\ref{fig:trajectory}B we drew an arrow originating from each grid vertex, pointing in the direction of the average angle formed by the line segments that passed within 2.4 units.  We set the arrow lengths to be inversely proportional to the $p$-values of the Rayleigh tests at each vertex.  Specifically, for each vertex we converted all of the angles of segments that passed within 2.4 units to unit vectors, and we set the arrow lengths at each vertex proportional to the length of the (circular) mean vector.  We also indicated any significant results ($p < 0.05$, corrected using the Benjamani-Hochberg procedure) by coloring the arrows in blue (darker blue denotes a lower $p$-value, i.e., a longer mean vector); all tests with $p \geq 0.05$ are displayed in gray and given a lower opacity value.

@@

\subsection*{Neurosynth decoding analyses}
\texttt{Neurosynth} parses a massive online database of over 14,000 neuroimaging studies and constructs meta-analysis images for over 13,000 psychology- and neuroscience-related terms, based on NIfTI images accompanying studies where those terms appear at a high frequency.  Given a novel image (tagged with its value type; e.g., $t$-, $F$- or $p$-statistics), \texttt{Neurosynth} returns a list of terms whose meta-analysis images are most similar.  Our permutation procedure yielded, for each of the two searchlight analyses, a voxelwise map of $z$-values.  These maps describe the extent to which each voxel \textit{specifically} reflected the temporal structure of the episode or individuals' recalls (i.e., relative to the null distributions of phase-shifted values). We inputted the two statistical maps described above to \texttt{Neurosynth} to create a list of the 10 most representative terms for each map.

>>>>

Our general approach draws inspiration from prior work aimed at elucidating the neural and behavioral underpinnings of how we process dynamic naturalistic experiences and remember them later.  Our approach to identifying neural responses to naturalistic stimuli (including experiences) entails building an explicit model of the stimulus dynamics and searching for brain regions whose responses are consistent with the model~\citep[also see][]{HuthEtal12, HuthEtal16}.  In prior work, a series of studies from Uri Hasson's group~\citep{LernEtal11, SimoEtal16, ChenEtal17, BaldEtal17, ZadbEtal17} have developed a clever alternative approach: rather than building an explicit stimulus model, these studies instead search for brain responses (while experiencing the stimulus) that are reliably similar across individuals.  So called \textit{inter-subject correlation} (ISC) and \textit{inter-subject functional connectivity} (ISFC) analyses effectively treat other people's brain responses to the stimulus as a ``model'' of how its features change over time~\citep[also see][]{SimoChan20}.  These purely brain-driven approaches are well suited to identifying which brain structures exhibit similar stimulus-driven responses across individuals.  Further, because neural response dynamics are observed data (rather than model approximations), such approaches do not require a detailed understanding of which stimulus properties or features might be driving the observed responses.  However, this also means that the specific stimulus features driving those responses are typically opaque to the researcher.  Our approach is complementary.  By explicitly modeling the stimulus dynamics, we are able to relate specific stimulus features to behavioral and neural dynamics.  However, when our model fails to accurately capture the stimulus dynamics that are truly driving behavioral and neural responses, our approach necessarily yields an incomplete characterization of the neural basis of the processes we are studying.

@@

The video stimulus was divided into 1,000 fine-grained ``time segments'' and annotated by an independent coder.  For each of these 1,000 annotations, the following information was recorded: a brief narrative description of what was happening, the location where the time segment took place, whether that location was indoors or outdoors, the names of all characters on-screen, the name(s) of the character(s) in focus in the shot, the name(s) of the character(s) currently speaking, the camera angle of the shot, a transcription of any text appearing on-screen, and whether or not there was music present in the background.  Each time segment was also tagged with its onset and offset time, in both seconds and TRs.

@@

\subsection*{Modeling the dynamic content of the episode and recall transcripts}
\subsubsection*{Topic modeling}
The input to the topic model we trained to characterize the dynamic content of the episode comprised 998 hand-generated annotations of short (mean: 2.96s) time segments spanning the video clip~(\citealp{ChenEtal17} generated 1000 annotations total; we removed two annotations referring to a break between the first and second scan sessions, during which no fMRI data were collected).  We concatenated the text for all of the annotated features within each segment, creating a ``bag of words'' describing its content and performed some minor preprocessing (e.g., stemming possessive nouns and removing punctuation).  We then re-organized the text descriptions into overlapping sliding windows spanning (up to) 50 annotations each.  In other words, we estimated the ``context'' for each annotated segment using the text descriptions of the preceding 25 annotations, the present annotations, and the following 24 annotations.  To model the context for annotations near the beginning of the episode (i.e., within 25 of the beginning or end), we created overlapping sliding windows that grew in size from one annotation to the full length.  We also tapered the sliding window lengths at the end of the episode, whereby time segments within fewer than 24 annotations of the end of the episode were assigned sliding windows that extended to the end of the episode.  This procedure ensured that each annotation's content was represented in the text corpus an equal number of times.
% ------ NOTE: ------
% need to refine explanation here a little bit
% - Paxton

We trained our model using these overlapping text samples with \texttt{scikit-learn}~\citep[version 0.19.1; ][]{PedrEtal11}, called from our high-dimensional visualization and text analysis software, \texttt{HyperTools}~\citep{HeusEtal18a}.  Specifically, we used the \texttt{CountVectorizer} class to transform the text from each window into a vector of word counts (using the union of all words across all annotations as the ``vocabulary,'' excluding English stop words); this yielded a number-of-windows by number-of-words \textit{word count} matrix.  We then used the \texttt{LatentDirichletAllocation} class (topics=100, method=`batch') to fit a topic model~\citep{BleiEtal03} to the word count matrix, yielding a number-of-windows (1047) by number-of-topics (100) \textit{topic proportions} matrix.  The topic proportions matrix describes the gradually evolving mix of topics (latent themes) present in each annotated time segment of the episode.  Next, we transformed the topic proportions matrix to match the 1976 fMRI volume acquisition times.  We assigned each topic vector to the timepoint (in seconds) midway between the beginning of the first annotation and the end of the last annotation in its corresponding sliding text window.  By doing so, we warped the linear temporal distance between consecutive topic vectors to align with the inconsistent temporal distance between consecutive annotations (whose durations varied greatly).  We then rescaled these timepoints to 1.5s TR units, and used linear interpolation to estimate a topic vector for each TR.  This resulted in a number-of-TRs (1976) by number-of-topics (100) matrix.

We created similar topic proportions matrices using hand-annotated transcripts of each participant's verbal recall of the episode~\citep[annotated by][]{ChenEtal17}.  We tokenized the transcript into a list of sentences, and then re-organized the list into overlapping sliding windows spanning (up to) 10 sentences each, analogously to how we parsed the episode annotations.  In turn, we transformed each window's sentences into a word count vector (using the same vocabulary as for the episode model), then used the topic model already trained on the episode scenes to compute the most probable topic proportions for each sliding window.  This yielded a number-of-windows (range: 83--312) by number-of-topics (100) topic proportions matrix for each participant.  These reflected the dynamic content of each participant's recalls.  Note: for details on how we selected the episode and recall window lengths and number of topics, see \textit{Supporting Information} and Figure~\topicopt.

@@

\paragraph{Probability of first recall (PFR).}  PFR curves~\citep{WelcBurn24, PostPhil65, AtkiShif68} reflect the probability that an item will be recalled first, as a function of its serial position during encoding. To carry out this analysis, we initialized a number-of-participants (17) by number-of-episode-events (30) matrix of zeros. Then, for each participant, we found the index of the episode event that was recalled first (i.e., the episode event whose topic vector was most strongly correlated with that of the first recall event) and filled in that index in the matrix with a 1.  Finally, we averaged over the rows of the matrix, resulting in a 1 by 30 array representing the proportion of participants that recalled an event first, as a function of the order of the event's appearance in the episode (Fig.~\ref{fig:list-learning}A).
% ------ NOTE: ------
% reiterate meaning of error ribbons in list-learning figure? (already noted in figure caption)
% - Paxton

@@

\paragraph{Serial position curve (SPC).} SPCs~\citep{Murd62a} reflect the proportion of participants that remember each item as a function of the item's serial positions during encoding. We initialized a number-of-participants (17) by number-of-episode-events (30) matrix of zeros. Then, for each recalled event, for each participant, we found the index of the episode event that the recalled event most closely matched (via the correlation between the events' topic vectors) and entered a 1 into that position in the matrix. This resulted in a matrix whose entries indicated whether or not each event was recalled by each participant (depending on whether the corresponding entires were set to one or zero).  Finally, we averaged over the rows of the matrix to yield a 1 by 30 array representing the proportion of participants that recalled each event as a function of the events' order appearance in the episode (Fig.~\ref{fig:list-learning}C).

@@

\subsubsection*{Visualizing the episode and recall topic trajectories}
We used the UMAP algorithm~\citep{McInEtal18} to project the 100-dimensional topic space onto a two-dimensional space for visualization (Figs.~\ref{fig:trajectory}, \ref{fig:topics}).  To ensure that all of the trajectories were projected onto the \textit{same} lower dimensional space, we computed the low-dimensional embedding on a ``stacked'' matrix created by vertically concatenating the events-by-topics topic proportions matrices for the episode, across-participants average recall and all 17 individual participants' recalls.  We then separated the rows of the result (a total-number-of-events by two matrix) back into individual matrices for the episode topic trajectory, across-participant average recall trajectory, and the trajectories for each individual participant's recalls (Fig.~\ref{fig:trajectory}).  This general approach for discovering a shared low-dimensional embedding for a collections of high-dimensional observations follows \cite{HeusEtal18a}.

@@

\subsubsection*{Estimating the consistency of flow through topic space across participants}
In Figure~\ref{fig:trajectory}B, we present an analysis aimed at characterizing locations in topic space that different participants move through in a consistent way (via their recall topic trajectories).  The two-dimensional topic space used in our visualizations (Fig.~\ref{fig:trajectory}) comprised a $60 \times 60$ (arbitrary units) square.  We tiled this space with a $50 \times 50$ grid of evenly spaced vertices, and defined a circular area centered on each vertex whose radius was two times the distance between adjacent vertices (i.e., 2.4 units).  For each vertex, we examined the set of line segments formed by connecting each pair successively recalled events, across all participants, that passed through this circle.  We computed the distribution of angles formed by those segments and the $x$-axis, and used a Rayleigh test to determine whether the distribution of angles was reliably ``peaked'' (i.e., consistent across all transitions that passed through that local portion of topic space).  To create Figure~\ref{fig:trajectory}B, we drew an arrow originating from each grid vertex, pointing in the direction of the average angle formed by the line segments that passed within 2.4 units.  We set the arrow lengths to be inversely proportional to the $p$-values of the Rayleigh tests at each vertex.  Specifically, for each vertex we converted all of the angles of segments that passed within 2.4 units to unit vectors, and we set the arrow lengths at each vertex proportional to the length of the (circular) mean vector.  We also indicated any significant results ($p < 0.05$, corrected using the Benjamani-Hochberg procedure) by coloring the arrows in blue (darker blue denotes a lower $p$-value, i.e., a longer mean vector); all tests with $p \geq 0.05$ are displayed in gray and given a lower opacity value.

@@

\subsection*{Neurosynth decoding analyses}
\texttt{Neurosynth} parses a massive online database of over 14,000 neuroimaging studies and constructs meta-analysis images for over 13,000 psychology- and neuroscience-related terms, based on NIfTI images accompanying studies where those terms appear at a high frequency.  Given a novel image (tagged with its value type; e.g., $z$-, $t$-, $F$- or $p$-statistics), \texttt{Neurosynth} returns a list of terms whose meta-analysis images are most similar.  Our permutation procedure yielded, for each of the two searchlight analyses, a voxelwise map of $z$-values.  These maps describe the extent to which each voxel \textit{specifically} reflected the temporal structure of the episode or individuals' recalls (i.e., relative to the null distributions of phase-shifted values). We inputted the two statistical maps described above to \texttt{Neurosynth} to create a list of the 10 most representative terms for each map.
